\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-10-07 paper.qmd
Changed the example dataset name

```{r}
#| label: fig-two-curvy-true-proj
#| fig-cap: "Two views of the true model (blue lines) in $2\\text{-}D$ projections from $7\\text{-}D$, for the two-curvy clusters data (grey points). The data is spread along the non-linear shaped clusters, and does not vary much from the curves. The **langevitour** software is used to view the data with a tour, and the full video is available at <>." 
#| fig-pos: H
#| fig-width: 10
#| fig-height: 5

@@

```{r}
#| label: fig-NLDR-two-curvy
#| echo: false
#| fig-cap: "Key steps for constructing the model on the UMAP layout ($k=2$): (a) data, (b) hexagon bins, (c) bin centroids, and (d) triangulated centroids. The two-curvy clusters data is shown."
#| fig-width: 12
#| fig-height: 4
#| out-width: 100%

@@

<!-- langevitour with pD model for two-curvy clusters -->
```{r}
#| label: best-fit-umap1
#| fig-cap: "Model in \\gD{} ($a_1 = 0.12$), on the layout and two views of the fit in projections from $7\\text{-}D$, for the two-curvy clusters data ($n =  2000$ and $p = 7$). One view shows the fitted model, and the other shows the fitted and true models alongside the data. The fitted model accurately represents the shape of the clusters but does not capture the edges of the structure. This illustrates a characteristic of UMAP, which compresses the data when transforming the data into \\gD{}. Video of the langevitour animations is available at <>."
#| fig-width: 10
#| fig-height: 8

@@

```{r}
#| echo: false
#| label: fig-bins-two-curvy
#| fig-pos: H
#| fig-cap: "Hexbin density plots of UMAP layout of the two-curvy clusters data, using three different bin inputs: (a) $b = 80 \\text{ } (10, \\text{ }8)$, (b) $b = 130 \\text{ } (13, \\text{ }10)$, and (c) $b = 391 \\text{ } (23, \\text{ }17)$. Color indicates standardized counts, dark indicating high count and light indicates low count. Using the MSE of the model fit in $7\\text{-}D$ helps decide on a useful choice of number of bins."
#| fig-width: 9
#| fig-height: 2

@@

The Delaunay triangulation will ensure that all centroids are connected into a triangular mesh. For some structures, like clustered data, or highly non-linear shapes, breaks in the mesh are meaningful. When separated clusters are present the mesh should be broken across the gaps. For non-linear structures like the S-curve, the mesh should run unbroken along the S, but there should be no edges connecting the top of the S directly to the bottom of the S. For these reasons it is necessary to remove edges from the mesh in some applications.

@@

```{r}
#| fig-cap: "Assessing which of the 5 NLDR layouts on the two-curvy clusters data is the better representation using MSE for varying binwidth ($a_1$). Colour used for the lines and points in the left plot and in the scatterplots represents NLDR layout (a-e). Layout d is universally poor. Layouts a that show two close clusters are universally suboptimal. Layout a is the best choice."
#| label: fig-two_curvy_clust-mse
#| fig-pos: H
#| out-height: 60%

@@

```{r}
#| label: best-fit-umap-tsne
#| fig-cap: "Best fit for two-curvy clusters data using UMAP (a1-a4) ($a_1$ = 0.09) and tSNE (b1-b4) ($a_1$ = 0.16) with the same number of non-empty bins ($m$ = 39). Both models show twist in $7\\text{-}D$ (a3, a4, b3, b4). Also, both NLDR methods are unable to accurately capture the width of the curvilinear clusters."
#| fig-width: 10
#| fig-height: 20
#| out-height: 90%

@@

It’s important to access the \gD{} layout and the generated model overlaid on data in \pD{} together to understand whether it fits the points everywhere, fits better in some places, or simply mismatches the pattern. Interactivity also helps in understanding the quirks that occur with different NLDR techniques (XXXXRefer the video after finalising the two-curvy clusters data).

>>>>

```{r}
#| label: fig-two-curvy-true-proj
#| fig-cap: "Two views of the true model (blue lines) in $2\\text{-}D$ projections from $7\\text{-}D$, for the two C-shaped clusters data (grey points). The data is spread along the non-linear shaped clusters, and does not vary much from the curves. The **langevitour** software is used to view the data with a tour, and the full video is available at <>." 
#| fig-pos: H
#| fig-width: 10
#| fig-height: 5

@@

```{r}
#| label: fig-NLDR-two-curvy
#| echo: false
#| fig-cap: "Key steps for constructing the model on the UMAP layout ($k=2$): (a) data, (b) hexagon bins, (c) bin centroids, and (d) triangulated centroids. The two C-shaped clusters data is shown."
#| fig-width: 12
#| fig-height: 4
#| out-width: 100%

@@

<!-- langevitour with pD model for two-curvy clusters -->
```{r}
#| label: best-fit-umap1
#| fig-cap: "Model in \\gD{} ($a_1 = 0.12$), on the layout and two views of the fit in projections from $7\\text{-}D$, for the two C-shaped clusters data ($n =  2000$ and $p = 7$). One view shows the fitted model, and the other shows the fitted and true models alongside the data. The fitted model accurately represents the shape of the clusters but does not capture the edges of the structure. This illustrates a characteristic of UMAP, which compresses the data when transforming the data into \\gD{}. Video of the langevitour animations is available at <>."
#| fig-width: 10
#| fig-height: 8

@@

```{r}
#| echo: false
#| label: fig-bins-two-curvy
#| fig-pos: H
#| fig-cap: "Hexbin density plots of UMAP layout of the two C-shaped clusters data, using three different bin inputs: (a) $b = 80 \\text{ } (10, \\text{ }8)$, (b) $b = 130 \\text{ } (13, \\text{ }10)$, and (c) $b = 391 \\text{ } (23, \\text{ }17)$. Color indicates standardized counts, dark indicating high count and light indicates low count. Using the MSE of the model fit in $7\\text{-}D$ helps decide on a useful choice of number of bins."
#| fig-width: 9
#| fig-height: 2

@@

The Delaunay triangulation will ensure that all centroids are connected into a triangular mesh. For some structures, like clustered data, or highly non-linear shapes, breaks in the mesh are meaningful. When separated clusters are present the mesh should be broken across the gaps. For non-linear structures like the C-shaped curvilinear, the mesh should run unbroken along the C, but there should be no edges connecting the top of the C directly to the bottom of the C. For these reasons it is necessary to remove edges from the mesh in some applications.

@@

```{r}
#| fig-cap: "Assessing which of the 5 NLDR layouts on the two C-shaped clusters data is the better representation using MSE for varying binwidth ($a_1$). Colour used for the lines and points in the left plot and in the scatterplots represents NLDR layout (a-e). Layout d is universally poor. Layouts a that show two close clusters are universally suboptimal. Layout a is the best choice."
#| label: fig-two_curvy_clust-mse
#| fig-pos: H
#| out-height: 60%

@@

```{r}
#| label: best-fit-umap-tsne
#| fig-cap: "Best fit for two C-shaped clusters data using UMAP (a1-a4) ($a_1$ = 0.09) and tSNE (b1-b4) ($a_1$ = 0.16) with the same number of non-empty bins ($m$ = 39). Both models show twist in $7\\text{-}D$ (a3, a4, b3, b4). Also, both NLDR methods are unable to accurately capture the width of the curvilinear clusters."
#| fig-width: 10
#| fig-height: 20
#| out-height: 90%

@@

It’s important to access the \gD{} layout and the generated model overlaid on data in \pD{} together to understand whether it fits the points everywhere, fits better in some places, or simply mismatches the pattern. Interactivity also helps in understanding the quirks that occur with different NLDR techniques (XXXXRefer the video after finalising the two C-shaped clusters data).

### 2024-10-07 paper.tex
Changed the example dataset name

\caption{\label{fig-two-curvy-true-proj}Two views of the true model
(blue lines) in \(2\text{-}D\) projections from \(7\text{-}D\), for the
two-curvy clusters data (grey points). The data is spread along the
non-linear shaped clusters, and does not vary much from the curves. The
\textbf{langevitour} software is used to view the data with a tour, and
the full video is available at \textless\textgreater.}

@@

\caption{\label{fig-NLDR-two-curvy}Key steps for constructing the model
on the UMAP layout (\(k=2\)): (a) data, (b) hexagon bins, (c) bin
centroids, and (d) triangulated centroids. The two-curvy clusters data
is shown.}

@@

\caption{Model in \gD{} (\(a_1 = 0.12\)), on the layout and two views of
the fit in projections from \(7\text{-}D\), for the two-curvy clusters
data (\(n =  2000\) and \(p = 7\)). One view shows the fitted model, and
the other shows the fitted and true models alongside the data. The
fitted model accurately represents the shape of the clusters but does
not capture the edges of the structure. This illustrates a
characteristic of UMAP, which compresses the data when transforming the
data into \gD{}. Video of the langevitour animations is available at
\textless\textgreater.}

@@

\caption{\label{fig-bins-two-curvy}Hexbin density plots of UMAP layout
of the two-curvy clusters data, using three different bin inputs: (a)
\(b = 80 \text{ } (10, \text{ }8)\), (b)
\(b = 130 \text{ } (13, \text{ }10)\), and (c)
\(b = 391 \text{ } (23, \text{ }17)\). Color indicates standardized
counts, dark indicating high count and light indicates low count. Using
the MSE of the model fit in \(7\text{-}D\) helps decide on a useful
choice of number of bins.}

@@

\caption{\label{fig-two_curvy_clust-mse}Assessing which of the 5 NLDR
layouts on the two-curvy clusters data is the better representation
using MSE for varying binwidth (\(a_1\)). Colour used for the lines and
points in the left plot and in the scatterplots represents NLDR layout
(a-e). Layout d is universally poor. Layouts a that show two close
clusters are universally suboptimal. Layout a is the best choice.}

@@

\caption{Best fit for two-curvy clusters data using UMAP (a1-a4)
(\(a_1\) = 0.09) and tSNE (b1-b4) (\(a_1\) = 0.16) with the same number
of non-empty bins (\(m\) = 39). Both models show twist in \(7\text{-}D\)
(a3, a4, b3, b4). Also, both NLDR methods are unable to accurately
capture the width of the curvilinear clusters.}

@@

It's important to access the \gD{} layout and the generated model
overlaid on data in \pD{} together to understand whether it fits the
points everywhere, fits better in some places, or simply mismatches the
pattern. Interactivity also helps in understanding the quirks that occur
with different NLDR techniques (XXXXRefer the video after finalising the
two-curvy clusters data).

>>>>

\caption{\label{fig-two-curvy-true-proj}Two views of the true model
(blue lines) in \(2\text{-}D\) projections from \(7\text{-}D\), for the
two C-shaped clusters data (grey points). The data is spread along the
non-linear shaped clusters, and does not vary much from the curves. The
\textbf{langevitour} software is used to view the data with a tour, and
the full video is available at \textless\textgreater.}

@@

\caption{\label{fig-NLDR-two-curvy}Key steps for constructing the model
on the UMAP layout (\(k=2\)): (a) data, (b) hexagon bins, (c) bin
centroids, and (d) triangulated centroids. The two C-shaped clusters
data is shown.}

@@

\caption{Model in \gD{} (\(a_1 = 0.12\)), on the layout and two views of
the fit in projections from \(7\text{-}D\), for the two C-shaped
clusters data (\(n =  2000\) and \(p = 7\)). One view shows the fitted
model, and the other shows the fitted and true models alongside the
data. The fitted model accurately represents the shape of the clusters
but does not capture the edges of the structure. This illustrates a
characteristic of UMAP, which compresses the data when transforming the
data into \gD{}. Video of the langevitour animations is available at
\textless\textgreater.}

@@

\caption{\label{fig-bins-two-curvy}Hexbin density plots of UMAP layout
of the two C-shaped clusters data, using three different bin inputs: (a)
\(b = 80 \text{ } (10, \text{ }8)\), (b)
\(b = 130 \text{ } (13, \text{ }10)\), and (c)
\(b = 391 \text{ } (23, \text{ }17)\). Color indicates standardized
counts, dark indicating high count and light indicates low count. Using
the MSE of the model fit in \(7\text{-}D\) helps decide on a useful
choice of number of bins.}

@@

\caption{\label{fig-two_curvy_clust-mse}Assessing which of the 5 NLDR
layouts on the two C-shaped clusters data is the better representation
using MSE for varying binwidth (\(a_1\)). Colour used for the lines and
points in the left plot and in the scatterplots represents NLDR layout
(a-e). Layout d is universally poor. Layouts a that show two close
clusters are universally suboptimal. Layout a is the best choice.}

@@

\caption{Best fit for two C-shaped clusters data using UMAP (a1-a4)
(\(a_1\) = 0.09) and tSNE (b1-b4) (\(a_1\) = 0.16) with the same number
of non-empty bins (\(m\) = 39). Both models show twist in \(7\text{-}D\)
(a3, a4, b3, b4). Also, both NLDR methods are unable to accurately
capture the width of the curvilinear clusters.}

@@

It's important to access the \gD{} layout and the generated model
overlaid on data in \pD{} together to understand whether it fits the
points everywhere, fits better in some places, or simply mismatches the
pattern. Interactivity also helps in understanding the quirks that occur
with different NLDR techniques (XXXXRefer the video after finalising the
two C-shaped clusters data).
